---
title: "Data Preparation Guide for SELC Analysts"
author: "V.Farrar"
output: html_document
---

```{r setup, include = FALSE}
library(tidyverse)
library(parameters)
```

In order to use the code included in the R Markdown files that generate reports, you will need to access the following variables in your institutional data, and format it following these guidelines.

# Dataset Preparation

## Dataset Format

Before running the code, ensure your data is formatted in the described SEISMIC format.

Course-level variables and student-level variables should be included in the same data table. Course-level variables should be present as separate columns. For instance, if students took multiple courses at an institution, there should be one row per student:course combination. Students that retook courses should thus have multiple rows for a single student:course combination that differ by term and grade. Student-level demographic data should be consistent across each student's rows. Note that this formatting differs from that of some other SEISMIC projects, where there are separate tables for course- and student-level variables.

Below is an example from the synthetic dataset. The first 3 columns, `crs_name`, `crs_section` and `n_section` are course level variables. The next columns are student-level variables, with demographic factors and then grade data as well.

**Note: The exact order in your dataset will not matter for running the code - only that all of the variables are present and named as described below.**

```{r load synthetic data, echo=FALSE, paged.print=TRUE}
synthdat <- read.csv(file = "~/Documents/GitHub/LAK-sandbox/SEISMIC_synthetic_data_2023-01-26.csv")

synthdat %>%
  select(crs_name, crs_section, n_section, st_id, female, firstgen,ethniccode_cat,international,transfer,lowincomeflag, numgrade, gpao) %>%
  arrange(crs_section) %>%
  head() %>%
  print_html()
```

For an example dataset, see: [Example dataset](https://docs.google.com/spreadsheets/d/14yD7tf09ZbpKUBF2KIkqPzPLpyKW9BeQyEd5zlR7EUM/edit?usp=sharing) (Google Doc; *open access*)

## Variables to Include

For the code to run, the following variables are required:

### Data Dictionary

<!-- example-table-LIST:START -->

+----------------+------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| variable       | type       | description / formatting notes                                                                                                                                                                                                                               |
+:===============+:===========+:=============================================================================================================================================================================================================================================================+
| st_id          | character  | unique identifier for each student (does not have to be actually linkable to students, can be random)                                                                                                                                                        |
+----------------+------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| female         | numeric    | student sex or gender (self-reported or registrar available)                                                                                                                                                                                                 |
|                |            |                                                                                                                                                                                                                                                              |
|                |            | 0: male                                                                                                                                                                                                                                                      |
|                |            |                                                                                                                                                                                                                                                              |
|                |            | 1: female                                                                                                                                                                                                                                                    |
|                |            |                                                                                                                                                                                                                                                              |
|                |            | "NA": other, third category                                                                                                                                                                                                                                  |
+----------------+------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| ethniccode_cat | numeric    | student race/ethnicity categories. Student self-identifies as:                                                                                                                                                                                               |
|                |            |                                                                                                                                                                                                                                                              |
|                |            | 0: White only                                                                                                                                                                                                                                                |
|                |            |                                                                                                                                                                                                                                                              |
|                |            | 1: includes *any or a mix* of Hispanic/Latino/Chicano, Black/African-American, American Indian/Indigenous, Alaska Native or Pacific Island                                                                                                                   |
|                |            |                                                                                                                                                                                                                                                              |
|                |            | 2: Asian/Asian-American only OR Asian/Asian-American and White.                                                                                                                                                                                              |
|                |            |                                                                                                                                                                                                                                                              |
|                |            | 3: Other                                                                                                                                                                                                                                                     |
|                |            |                                                                                                                                                                                                                                                              |
|                |            | NA: Missing or not available                                                                                                                                                                                                                                 |
+----------------+------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| firstgen       | numeric    | Parental education/ First-Generation college going status.                                                                                                                                                                                                   |
|                |            |                                                                                                                                                                                                                                                              |
|                |            | 0: one or more parent/guardian(s) has earned a Bachelor's degree or higher                                                                                                                                                                                   |
|                |            |                                                                                                                                                                                                                                                              |
|                |            | 1: no parent/guardian has earned a Bachelor's degree or higher                                                                                                                                                                                               |
+----------------+------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| lowincomeflag  | numeric    | Low-Income status.                                                                                                                                                                                                                                           |
|                |            |                                                                                                                                                                                                                                                              |
|                |            | 0: Student or family income not flagged as low-income.                                                                                                                                                                                                       |
|                |            |                                                                                                                                                                                                                                                              |
|                |            | 1: Student is flagged as low-income based on family household income and household size using 185% of the U.S. poverty line. Use Pell eligible if available; if not: \$25,100\*1.85 = \$46,435 (185% of U.S. poverty line for household size of 4; \$25,100) |
+----------------+------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| transfer       | numeric    | Transfer status from another institution (two-year, such as junior/community colleges, or four-year) \                                                                                                                                                       |
|                |            |                                                                                                                                                                                                                                                              |
|                |            | 0: student not flagged as transfer; student entered institution as a first-year / freshman; prior institution was high-school                                                                                                                                |
|                |            |                                                                                                                                                                                                                                                              |
|                |            | 1: student flagged as transfer; students' prior institution was another higher education institution (two-year or four-year)                                                                                                                                 |
+----------------+------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| international  | numeric    | International / Non-resident alien status.                                                                                                                                                                                                                   |
|                |            |                                                                                                                                                                                                                                                              |
|                |            | 0: U.S. Citizen / permanent resident                                                                                                                                                                                                                         |
|                |            |                                                                                                                                                                                                                                                              |
|                |            | 1: Non-resident alien / international student.                                                                                                                                                                                                               |
+----------------+------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| crs_name       | character  | Name of the course of interest. Recommend course catalog numbers (e.g. BIO100) instead of full names ("Fundamentals of Biology")                                                                                                                             |
+----------------+------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| crs_term       | character  | Academic term the course was offered. Any format can be used here, especially if this will link with the notation in your institutional database.                                                                                                            |
|                |            |                                                                                                                                                                                                                                                              |
|                |            | *Note: specific course years, semesters/quarters, and sections are stored in other variables.*                                                                                                                                                               |
+----------------+------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| crs_year       | numeric    | Calendar year the course was offered. Enter in "YYYY" format.                                                                                                                                                                                                |
+----------------+------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| crs_semq       | numeric    | Semester or quarter the course was offered. Use the following notation:                                                                                                                                                                                      |
|                |            |                                                                                                                                                                                                                                                              |
|                |            | 01: Spring or Winter semester; Winter quarter (UC)                                                                                                                                                                                                           |
|                |            |                                                                                                                                                                                                                                                              |
|                |            | 03: Spring quarter (UC)                                                                                                                                                                                                                                      |
|                |            |                                                                                                                                                                                                                                                              |
|                |            | 05: Summer session, or Summer session 1/I                                                                                                                                                                                                                    |
|                |            |                                                                                                                                                                                                                                                              |
|                |            | 07: Summer session 2/II                                                                                                                                                                                                                                      |
|                |            |                                                                                                                                                                                                                                                              |
|                |            | 10: Fall semester or Fall quarter (UCs)                                                                                                                                                                                                                      |
+----------------+------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| crs_section    | numeric    | Section of the course within a term. If more than one section was offered of the course per term, use numeric codes following your institutions notation (e.g. "01", "02", "03")                                                                             |
|                |            |                                                                                                                                                                                                                                                              |
|                |            | Note: If only one section was offered, please default to "01".                                                                                                                                                                                               |
+----------------+------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| numgrade       | numeric    | Student's final course grade in the course listed in `crs_name` and term in `crs_term`\                                                                                                                                                                      |
|                |            | Please convert to a numeric grade using your institution's grading conventions. ([Example scale from UC Davis here](https://registrar.ucdavis.edu/records/transcripts/calculate-gpa)) \                                                                      |
|                |            | *Note: Typically ranges from 0-4.0*                                                                                                                                                                                                                          |
+----------------+------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| gpao           | numeric    | Grade point average omitting the course of interest (GPAO). Student's cumulative grade point average at the end of the term listed in `crs_term`, minus the grade in the course of interest (`crs_name`).                                                    |
+----------------+------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+


# Check Your Data
