---
title: "SEISMIC Equity Measures"
subtitle: "Level 1"
output:
  rmdformats::readthedown
css:
  custom.css
date: "2023-03-14"

---

*(This R Markdown is a demo example based upon Questions (version 3) in [this Google Doc](https://docs.google.com/document/d/1TceXPFyIDZ0ZC06XFtYz6fWkgvnNYbpq7F11voeGLFA/edit#heading=h.h9qcp91x0k23)).*

```{r setup, include=FALSE}
#hide code, warnings, and messages by default
knitr::opts_chunk$set(include = FALSE, warnings = FALSE, messages = FALSE)
library(tidyverse)
library(rmdformats) #output format: readthedown
library(kableExtra) #pretty html tables
```

```{r load data, include = FALSE}
#load data 
#change this file name to the source of your properly formatted data file 
dat <- read.csv(file = "~/Documents/GitHub/LAK-sandbox/SEISMIC_synthetic_data_2023-01-26.csv")
```

# Who is in this course?

What is the makeup of the course by

-   race/ethnicity

-   gender

-   first-generation status

-   socioeconomic status

-   transfer status

-   STEM major

-   X major

-   in-state

-   international

-   military

```{r demographics bar plot}
#bar plot with percent of all students in the class by demographics

```

## How have these demographics changed over time?
```{r demographics over time}
#different identities with the percent over time
  #could be facet grid or not
```

# Grades in this course?

## How do course outcomes compare for different student groups?
```{r raw grades by identity}
#different identities with raw grades on average across identities 
```

## How does oveall prior performance at the university (i.e., GPAO, prior GPA) compare to student outcomes in my course?

[grade anomaly explainer here]

```{r grade anomaly by identity}
#different identities with average grade anomaly across identities 
```

## How do course outcomes differ across the spectrum of systemic advantges students may have access to in higher education?

### SAI
[SAI explainer here]

```{r raw grades by SAI}
#raw grades by SAI
```

```{r grade anomaly by SAI}
#grade anomaly by SAI
```

```{r R coding resources}
#R coding resources used

```

